Sample Size Determination: Posterior Distributions Proximity

### 2024-05-14 paper/main.tex
changed style file

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.21 of 2022/01/12
%
\documentclass[runningheads]{llncs}
\input{preamble}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
\usepackage{color}
\renewcommand\UrlFont{\color{blue}\rmfamily}
\urlstyle{rm}
%
\begin{document}
%
\title{Sample Size Determination:\\Posterior Distributions Proximity}
%
\titlerunning{Sample Size Determination: Posterior Distributions Proximity}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
%\author{Nikita Kiselev\inst{1} \and Andrey Grabovoy\inst{1}}
\author{Nikita Kiselev \and Andrey Grabovoy}
%
\authorrunning{N. Kiselev, A. Grabovoy}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Moscow Institute of Physics and Technology, Moscow, Russia}
%
\maketitle % typeset the header of the contribution
%
\begin{abstract}
The paper investigates the problem of estimating a sufficient sample size. The issue of determining a sufficient sample size without specifying a statistical hypothesis about the distribution of model parameters is considered. Two approaches to determining a sufficient sample size based on the proximity of posterior distributions of model parameters on similar subsets are suggested. The correctness of the presented approaches is proven in the model with normal posterior distribution. A theorem about moments of the limit posterior distribution of parameters in a linear regression model is proven. A computational experiment is conducted to analyze the properties of the proposed methods.

\keywords{Sufficient sample size \and Posterior distributions proximity \and Bayesian inference \and Linear regression.}
\end{abstract}
%
%
%
\section{Introduction}
The task of supervised machine learning involves selecting a predictive model from a parametric family. This choice is usually based on certain statistical hypotheses, such as maximizing a quality functional. A model that satisfies these statistical hypotheses is called an \textit{adequate} model \cite{bies2006genetic,cawley2010over,raschka2018model}.

@@

\section{Computational experiment}
This section provides an empirical study of the proposed methods. Experiments were conducted on synthetic data and dataset Liver Disorders from \cite{UCI}.

@@

@@

%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
\clearpage
\bibliographystyle{splncs04}
\bibliography{references}
%

\clearpage
\appendix

\section{Proof of Theorem \ref{theorem1}}
\begin{proof}
The Kullback-Leibler divergence for a pair of normal posterior distributions has the form
    \[ D_{\text{KL}}\left( p_k \| p_{k+1} \right) = \dfrac{1}{2} \left( \mathrm{tr}\left( \mathbf{\Sigma}_{k+1}^{-1} \mathbf{\Sigma}_k \right) + (\mathbf{m}_{k+1} - \mathbf{m}_k)\T \mathbf{\Sigma}_{k+1}^{-1} (\mathbf{m}_{k+1} - \mathbf{m}_k) - n + \log{\left( \dfrac{\det \mathbf{\Sigma}_{k+1}}{\det \mathbf{\Sigma}_{k}} \right)} \right). \]
    Let's express $\mathbf{\Sigma}_{k+1}$ as $\mathbf{\Sigma}_{k+1} = \mathbf{\Sigma}_k + \Delta\mathbf{\Sigma}$. Let's consider each term separately.
    \[ \mathrm{tr}\left( \mathbf{\Sigma}_{k+1}^{-1} \mathbf{\Sigma}_k \right) = \mathrm{tr}\left(\left(\mathbf{\Sigma}_k + \Delta \mathbf{\Sigma} \right)^{-1} \mathbf{\Sigma}_k \right) \to \mathrm{tr}\mathbf{I}_n=n\text{ as } \| \Delta \mathbf{\Sigma} \|_F \to 0, \]
    \[ \left| (\mathbf{m}_{k+1} - \mathbf{m}_k)\T\mathbf{\Sigma}_{k+1}^{-1} (\mathbf{m}_{k+1} - \mathbf{m}_k) \right| \leqslant\| \mathbf{m}_{k+1} -\mathbf{m}_k\|_2^2\|\mathbf{\Sigma}_{k+1}^{-1} \|_2 \ to 0 \text{ as } \| \mathbf{m}_{k+1} - \mathbf{m}_k\|_2 \to 0, \]
    \[ \log{\left( \dfrac{\det \mathbf{\Sigma}_{k+1}}{\det \mathbf{\Sigma}_{k}} \right)} = \log{\left( \dfrac{\det \left( \mathbf{\Sigma}_k + \Delta \mathbf{\Sigma} \right)}{\det\mathbf{\Sigma}_{k}} \right)} \to \log \det\mathbf{I}_n = \log 1 = 0 \text{ as } \| \Delta \mathbf{\Sigma} \|_F\to 0, \]
    from where we have the required.
\end{proof}

\section{Proof of Theorem \ref{theorem2}}
\begin{proof}
Let's use the s-score expression for a pair of normal posterior distributions from \cite{Aduenko2017}:
\[\text{s-score}(p_k, p_{k+1}) = \exp{\left( -\dfrac{1}{2} (\mathbf{m}_{k+1} - \mathbf{m}_k)\T \left( \mathbf{\Sigma}_k + \mathbf{\Sigma}_{k+1} \right)^{-1} (\mathbf{m}_{k+1} - \mathbf{m}_k) \right)}. \]
Because
    \[ \left| (\mathbf{m}_{k+1} - \mathbf{m}_k)\T \left( \mathbf{\Sigma}_k + \mathbf{\Sigma}_{k+1} \right)^{-1} (\mathbf{m}_{k+1} - \mathbf{m}_k) \right| \leqslant \| \mathbf{m}_{k+1} - \mathbf{m}_k \|_2^2 \| \left( \mathbf{\Sigma}_k + \mathbf{\Sigma}_{k+1} \right)^{-1} \|_2 \to 0 \]
    if $\|\mathbf{m}_{k+1} - \mathbf{m}_k\|_2\to 0$, then the value of the quadratic form inside the exponent tends to zero. Therefore, $\text{s-score}(p_k, p_{k+1}) \to 1$ as $\|\mathbf{m}_{k+1} - \mathbf{m}_k\|_2\to 0$.
\end{proof}

@@

>>>>

%Version 3 October 2023
% See section 11 of the User Manual for version history
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%\documentclass[referee,sn-basic]{sn-jnl}% referee option is meant for double line spacing

%%=======================================================%%
%% to print line numbers in the margin use lineno option %%
%%=======================================================%%

%%\documentclass[lineno,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style

%%======================================================%%
%% to compile with pdflatex/xelatex use pdflatex option %%
%%======================================================%%

%%\documentclass[pdflatex,sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style

%%Note: the following reference styles support Namedate and Numbered referencing. By default the style follows the most common style. To switch between the options you can add or remove Numbered in the optional parenthesis. 
%%The option is available for: sn-basic.bst, sn-vancouver.bst, sn-chicago.bst%

%%\documentclass[sn-nature]{sn-jnl}% Style for submissions to Nature Portfolio journals
%%\documentclass[sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
\documentclass[sn-mathphys-num]{sn-jnl}% Math and Physical Sciences Numbered Reference Style 
%%\documentclass[sn-mathphys-ay]{sn-jnl}% Math and Physical Sciences Author Year Reference Style
%%\documentclass[sn-aps]{sn-jnl}% American Physical Society (APS) Reference Style
%%\documentclass[sn-vancouver,Numbered]{sn-jnl}% Vancouver Reference Style
%%\documentclass[sn-apa]{sn-jnl}% APA Reference Style 
%%\documentclass[sn-chicago]{sn-jnl}% Chicago-based Humanities Reference Style

%%%% Standard Packages
%%<additional latex packages if required can be included here>

\input{preamble}

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
%%%%

%%%%%=============================================================================%%%%
%%%%  Remarks: This template is provided to aid authors with the preparation
%%%%  of original research articles intended for submission to journals published 
%%%%  by Springer Nature. The guidance has been prepared in partnership with 
%%%%  production teams to conform to Springer Nature technical requirements. 
%%%%  Editorial and presentation requirements differ among journal portfolios and 
%%%%  research disciplines. You may find sections in this template are irrelevant 
%%%%  to your work and are empowered to omit any such section if allowed by the 
%%%%  journal you intend to submit to. The submission guidelines and policies 
%%%%  of the journal take precedence. A detailed User Manual is available in the 
%%%%  template package for technical guidance.
%%%%%=============================================================================%%%%

%% as per the requirement new theorem styles can be included as shown below
%\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}% 
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

%theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%

%\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}%

\raggedbottom
%%\unnumbered% uncomment this for unnumbered level heads

\begin{document}

\title[Sample Size Determination: Posterior Distributions Proximity]{Sample Size Determination: Posterior Distributions Proximity}

%%=============================================================%%
%% GivenName	-> \fnm{Joergen W.}
%% Particle	-> \spfx{van der} -> surname prefix
%% FamilyName	-> \sur{Ploeg}
%% Suffix	-> \sfx{IV}
%% \author*[1,2]{\fnm{Joergen W.} \spfx{van der} \sur{Ploeg} 
%%  \sfx{IV}}\email{iauthor@gmail.com}
%%=============================================================%%

\author*[1]{\fnm{Nikita} \sur{Kiselev}}\email{kiselev.ns@phystech.edu}

\author[1]{\fnm{Andrey} \sur{Grabovoy}}\email{grabovoy.av@phystech.edu}

\affil[1]{\orgname{Moscow Institute of Physics and Technology}, \orgaddress{\city{Dolgoprudny}, \country{Russia}}}

%%==================================%%
%% Sample for unstructured abstract %%
%%==================================%%

\abstract{The issue of sample size determination is crucial for constructing an effective machine learning model. However, the existing methods for determining a sufficient sample size are either not strictly proven, or relate to the specific statistical hypothesis about the distribution of model parameters. In this paper we present two approaches based on the proximity of posterior distributions of model parameters on similar subsamples. We show that these two methods are valid for the model with normal posterior distribution of parameters. Computational experiments demonstrate the convergence of the proposed functions as the sample size increases.}

\keywords{Sufficient sample size, Posterior distributions proximity, Normal posterior distribution, Linear regression}

%%\pacs[JEL Classification]{D8, H51}

%%\pacs[MSC Classification]{35A01, 65L10, 65L12, 65L20, 65L70}

\maketitle

\section{Introduction}
\label{sec:introduction}
The task of supervised machine learning involves selecting a predictive model from a parametric family. This choice is usually based on certain statistical hypotheses, such as maximizing a quality functional. A model that satisfies these statistical hypotheses is called an \textit{adequate} model \cite{bies2006genetic,cawley2010over,raschka2018model}.

@@

\section{Computational experiment}

This section provides an empirical study of the proposed methods. Experiments were conducted on synthetic data and dataset Liver Disorders from \cite{UCI}.

@@

To compare the performance of the proposed method on different datasets, samples have been chosen from the open repository \cite{UCI}. The detailed information about each dataset, the number of observations and number of features, are provided in Table~\ref{table}. For demonstration purposes, a value of the hyperparameter $\varepsilon$ has been selected, at which the value of the target function, either $KL(k)$ or $S(k)$, decreases by half. The corresponding results are in Table~\ref{table}. Omissions mean that the initial sample size is not sufficient.

\begin{table}[ht]
    \centering
    \caption{Datasets with regression task (omissions mean that the initial sample size is not sufficient)}\label{table}
    \begin{tabular}{ccccc}
    \toprule
    Dataset name & Objects $m$ & Features $n$ & KL & S \\
    \midrule
    Abalone & 4177 & 8 & 3751 & 3751\\
    Auto MPG & 392 & 8 & 55 & --- \\
    Automobile & 159 & 25 & 152 & --- \\
    Liver Disorders & 345 & 6 & 282 & --- \\
    Servo & 167 & 4 & 138 & 160 \\
    Forest fires & 517 & 12 & 487 & --- \\
    Wine Quality & 6497 & 12 & --- & --- \\
    Energy Efficiency & 768 & 9 & --- & --- \\
    Student Performance & 649 & 32 & --- & --- \\
    Facebook Metrics & 495 & 18 & 379 & 475  \\
    Real Estate Valuation & 414 & 7 & --- & --- \\
    Heart Failure Clinical Records & 299 & 12 & 258 & 293 \\
    Bone marrow transplant: children & 142 & 36 & 109 & --- \\
    \botrule
    \end{tabular}
\end{table}

@@

\begin{appendices}

\section{Proof of Theorem \ref{theorem1}}
\begin{proof}
The Kullback-Leibler divergence for a pair of normal posterior distributions has the form
    \[ D_{\text{KL}}\left( p_k \| p_{k+1} \right) = \dfrac{1}{2} \left( \mathrm{tr}\left( \mathbf{\Sigma}_{k+1}^{-1} \mathbf{\Sigma}_k \right) + (\mathbf{m}_{k+1} - \mathbf{m}_k)\T \mathbf{\Sigma}_{k+1}^{-1} (\mathbf{m}_{k+1} - \mathbf{m}_k) - n + \log{\left( \dfrac{\det \mathbf{\Sigma}_{k+1}}{\det \mathbf{\Sigma}_{k}} \right)} \right). \]
    Let's express $\mathbf{\Sigma}_{k+1}$ as $\mathbf{\Sigma}_{k+1} = \mathbf{\Sigma}_k + \Delta\mathbf{\Sigma}$. Let's consider each term separately.
    \[ \mathrm{tr}\left( \mathbf{\Sigma}_{k+1}^{-1} \mathbf{\Sigma}_k \right) = \mathrm{tr}\left(\left(\mathbf{\Sigma}_k + \Delta \mathbf{\Sigma} \right)^{-1} \mathbf{\Sigma}_k \right) \to \mathrm{tr}\mathbf{I}_n=n\text{ as } \| \Delta \mathbf{\Sigma} \|_F \to 0, \]
    \[ \left| (\mathbf{m}_{k+1} - \mathbf{m}_k)\T\mathbf{\Sigma}_{k+1}^{-1} (\mathbf{m}_{k+1} - \mathbf{m}_k) \right| \leqslant\| \mathbf{m}_{k+1} -\mathbf{m}_k\|_2^2\|\mathbf{\Sigma}_{k+1}^{-1} \|_2 \to 0 \text{ as } \| \mathbf{m}_{k+1} - \mathbf{m}_k\|_2 \to 0, \]
    \[ \log{\left( \dfrac{\det \mathbf{\Sigma}_{k+1}}{\det \mathbf{\Sigma}_{k}} \right)} = \log{\left( \dfrac{\det \left( \mathbf{\Sigma}_k + \Delta \mathbf{\Sigma} \right)}{\det\mathbf{\Sigma}_{k}} \right)} \to \log \det\mathbf{I}_n = \log 1 = 0 \text{ as } \| \Delta \mathbf{\Sigma} \|_F\to 0, \]
    from where we have the required.
\end{proof}

\section{Proof of Theorem \ref{theorem2}}
\begin{proof}
Let's use the s-score expression for a pair of normal posterior distributions from~\cite{Aduenko2017}:
\[\text{s-score}(p_k, p_{k+1}) = \exp{\left( -\dfrac{1}{2} (\mathbf{m}_{k+1} - \mathbf{m}_k)\T \left( \mathbf{\Sigma}_k + \mathbf{\Sigma}_{k+1} \right)^{-1} (\mathbf{m}_{k+1} - \mathbf{m}_k) \right)}. \]
Because
    \[ \left| (\mathbf{m}_{k+1} - \mathbf{m}_k)\T \left( \mathbf{\Sigma}_k + \mathbf{\Sigma}_{k+1} \right)^{-1} (\mathbf{m}_{k+1} - \mathbf{m}_k) \right| \leqslant \| \mathbf{m}_{k+1} - \mathbf{m}_k \|_2^2 \| \left( \mathbf{\Sigma}_k + \mathbf{\Sigma}_{k+1} \right)^{-1} \|_2 \to 0 \]
    if $\|\mathbf{m}_{k+1} - \mathbf{m}_k\|_2\to 0$, then the value of the quadratic form inside the exponent tends to zero. Therefore, $\text{s-score}(p_k, p_{k+1}) \to 1$ as $\|\mathbf{m}_{k+1} - \mathbf{m}_k\|_2\to 0$.
\end{proof}

@@

\end{appendices}

\bibliography{references}% common bib file
%% if required, the content of .bbl file can be included here once bbl is generated
%%\input sn-article.bbl
